---
title: "Basic Statistical Analysis of Severe Weather Events in the United States"
output: 
  html_document:
    keep_md: true
---

## Synopsis

This study uses NOAA Storm Database's data to find out what types of events are most harmful to population's health and the types that cause most economical damage across the United States. Main findings are that tornadoes are most damaging health-wise, while floods have greatest economical impact. Also events that have high health effect are less severe on property damages, and vice-versa. The study is an assignment for course Reproducible Research by Roger D. Peng, Jeff Leek and Brian Caffo (Johns Hopkins University / Coursera).

## Data Processing

Data was downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2 at 2015-03-12, uncompressed with bzip2 and analysis was carried out with R by using following libraries.

```{r}
library(data.table)
library(ggplot2)
library(reshape2)
library(plyr)
library(grid)
```

To answer the question about effect on populations health, the data is aggregated by event types by calculating sums of fatalities and injuries.
```{r cache=TRUE}
csv <- read.csv(file = 'repdata_data_StormData.csv')
fatalitiesAndInjuriesSumByEventType <- aggregate(. ~ EVTYPE, data=csv[,c('EVTYPE', 'FATALITIES', 'INJURIES')], FUN=sum)
```

Because the number of different event types is large, only ten most harmful ones fatality- and injury-wise are visualised. The top ten is collected separately for both health impacts, because their numbers don't always correlate with each other.

```{r cache=TRUE}
pickTopTen <- function(data, columnIndex) {
  head(data[order(-data[,c(columnIndex)]),], 10)
}
topFatalitiesSumByEventType <- pickTopTen(fatalitiesAndInjuriesSumByEventType, 2)
topInjuriesSumByEventType <- pickTopTen(fatalitiesAndInjuriesSumByEventType, 3)
```

While investigating events' economical impact, similar aggregation is carried over total property and crop damage expenditures. The economical impact is in original data in two parts (numeric and exponent), and following also multiplies damages by the exponent.

```{r cache=TRUE}
csv$PROPDMGEXP <- revalue(csv$PROPDMGEXP, c("K"=1000, "M"=1000000, "B"=1000000000), warn_missing = FALSE)
csv$CROPDMGEXP <- revalue(csv$CROPDMGEXP, c("K"=1000, "M"=1000000, "B"=1000000000), warn_missing = FALSE)
suppressWarnings(csv$PROPDMG <- csv$PROPDMG * as.integer(as.character(csv$PROPDMGEXP)))
suppressWarnings(csv$CROPDMG <- csv$CROPDMG * as.integer(as.character(csv$CROPDMGEXP)))
damagesSumByEventType <- aggregate(. ~ EVTYPE, data=csv[,c('EVTYPE', 'PROPDMG', 'CROPDMG')], FUN=sum)
```

Top ten event types by economical damages are then sorted by sum of the damage types.

```{r cache=TRUE}
damagesSumByEventType$SUM <- damagesSumByEventType$PROPDMG + damagesSumByEventType$CROPDMG
topDamagesSumByEventType <- pickTopTen(damagesSumByEventType, 4)
```

Finally the helper (combined sum) variable is removed, the data is made narrow (long) so it is suitable for ggplot2's stacked bar chart, and also damage types are renamed.

```{r cache=TRUE}
topDamagesSumByEventType <- topDamagesSumByEventType[, c('EVTYPE', 'PROPDMG', 'CROPDMG')]
topDamagesSumByEventType <- melt(topDamagesSumByEventType, id=c("EVTYPE")) 
levels(topDamagesSumByEventType$variable)[levels(topDamagesSumByEventType$variable)=="PROPDMG"] <- "Property"
levels(topDamagesSumByEventType$variable)[levels(topDamagesSumByEventType$variable)=="CROPDMG"] <- "Crop"
```

## Results

### Personnel Damages

Following plot by total health impact shows that tornadoes are clearly the major cause of deaths. However, also excessive heat causes high number of fatalities, about one third compared to tornadoes. 


```{r}
fatalitiesSumPlot <- ggplot(topFatalitiesSumByEventType[,c('EVTYPE', 'FATALITIES')], aes(x=EVTYPE, y=FATALITIES)) + geom_bar(stat="identity") + coord_flip() + ylab('Fatalities (sum)') + xlab('') + ggtitle('Sum of fatalities by event type')
injuriesSumPlot <- ggplot(topInjuriesSumByEventType[,c('EVTYPE', 'INJURIES')], aes(x=EVTYPE, y=INJURIES)) + geom_bar(stat="identity")  + coord_flip() + ylab('Injuries (sum)') + xlab('') + ggtitle('Sum of injuries by event type')
layout <- matrix(seq(1, 2), ncol = 1, nrow = 2)
grid.newpage()
pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
print(fatalitiesSumPlot, vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
print(injuriesSumPlot, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
```

It would be interesting to investigate severity of the injuries: for example do tornadoes cause mostly light and easily treatable injuries, or more severe ones with significant economical impact in addition to the effect they have on victims' quality of life.

### Damages

Damages are investigated by plotting sum of combined property and crop damages by top ten event types.

```{r}
damagesSumPlot <- ggplot(topDamagesSumByEventType, aes(x=factor(EVTYPE), y=value, fill=variable)) + geom_bar(stat="identity") + coord_flip() + ylab('Cost (USD)') + xlab('') + scale_fill_discrete('Damage type')  + ggtitle('Damages by event type')
print(damagesSumPlot)
```

Floods cause most of damages, and compared to that tornadoes have a small economical impact. The differing economical and health effects could be caused by events' nature: floods are generally slower and leave more time to evacuate people, while tornadoes are sudden and violent events.

The statistics could be analysed further by location of the events, for example do floods repeatedly concentrate on single areas and whether it would be feasible to prevent damages by building flood barries or safe overflow areas. It is also worth noting that, except for river floods and ice storms, crop damages are small compared to property damages.